---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rdataset

<!-- badges: start -->
[![R-CMD-check](https://github.com/SchlossLab/rdataset/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/SchlossLab/rdataset/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/SchlossLab/rdataset/graph/badge.svg)](https://app.codecov.io/gh/SchlossLab/rdataset)
<!-- badges: end -->

The 'rdataset' package stores the data associated with your microbial DNA anaylsis. This includes nucleotide sequences, abundance, sample and treatment assignments, taxonomic classifications, sequence bin assignments, 
metadata, trees and various reports. It is designed to facilitate data analysis across multiple R packages with utility functions to read and write from [mothur](https://mothur.org), mothur2, [qiime2](https://qiime2.org), [dada2](https://benjjneb.github.io/dada2/) and [phyloseq](https://www.bioconductor.org/packages/release/bioc/html/phyloseq.html).

## Installation

You can install the CRAN version with:

```{r, eval = FALSE}
install.packages("rdataset")
```

You can install the development version of rdataset from [GitHub](https://github.com/rdataset) with:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("rdataset")
```

## Data Tranfer Functions

- 'read_mothur()' : Create a 'dataset' object from mothur's output files.
- 'read_dada2()' : Create a 'dataset' object from dada2's outputs.
- 'read_qiime2()' : Create a 'dataset' object from qiime2's .qza output files.
- 'read_phyloseq()' : Create a 'dataset' object from phyloseq's outputs.

## Get started

See the [Getting Started](TODO) 
page to get started.

## Getting help

If you encounter an issue, please file an issue on  [GitHub](https://github.com/rdataset/issues). Please include a minimal 
reproducible example with your issue.

## Contributing

Is there a feature you'd like to see included, please let us know! Pull requests
are welcome on [GitHub](https://github.com/rdataset/pulls). 
