---
title: "Data Transfers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Transfers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(strollur)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The *strollur* package stores the data associated with your microbial DNA analysis. This tutorial will explain how to save, load, copy, export, and import your dataset object. If you haven't reviewed the [Getting Started](Getting-Started.html) tuturial, we recommend you start there. 

Let's use the `miseq_sop_example()` function to create a dataset object from the [Miseq SOP Example](https://mothur.org/wiki/miseq_sop/). 

```{r}
miseq <- miseq_sop_example()
miseq
```

## Saving and Loading

The strollur package has a function to save a dataset object as an *.rds* file, `save_dataset()`, and a function to create a dataset from an *.rds* file, `load_dataset()`. Let's use the miseq data object to learn how to do that.


```{r}
save_dataset(miseq, file = "miseq_sop.rds")

miseq_from_rds <- load_dataset(file = "miseq_sop.rds")
miseq_from_rds
unlink("miseq_sop.rds")
```

We can see that the summaries of miseq and miseq_from_rds are identical. Let's modify miseq_from_rds to verify they are not referring to the same object. We will add clusters created by [mothur](https://mothur.org) using [vsearch's](https://github.com/torognes/vsearch) distance-based greedy clustering (dgc) algorithm. 

```{r}
dgc_data <- read_mothur_list(list = strollur_example("final.dgc.list.gz"))

assign(miseq_from_rds, table = dgc_data, bin_type = "dgc")
miseq_from_rds
miseq
```

We can see from the summary that 361 'dgc' bins were added to miseq_from_rds and not to miseq.

## Export and Import

The *.rds* file is in binary format and is not human readable. You can use the `export_dataset()` to see a human readable form of the raw data stored in the dataset. Let's export *miseq* and look at the table created.

```{r}
table <- export_dataset(miseq)
str(table)
```

Similarly to `load_dataset()`, you can use the `import_dataset()` function to create a new dataset object from the exported table. 

```{r}
miseq_import <- import_dataset(table = table)
miseq_import
```

Again, we can see that the summary of miseq_import is identical to the summary of miseq.

## Copy

Lastly, you can make a deep copy of your dataset using the `copy_dataset()` function. Note, if you use an assignment operator to copy it's a shallow copy. The dataset object is an R6 object to keep the memory usage low. First let's learn how to use the `copy_dataset()` function, then we will take a closer look at how deep and shallow copying differ.

```{r}
miseq_deep_copy <- copy_dataset(miseq)

miseq_shallow_copy <- miseq
```

Let's add the dgc_data to miseq_shallow_copy and then compare miseq, miseq_deep_copy, and mise_shallow_copy.

```{r}
assign(miseq_shallow_copy, table = dgc_data, bin_type = "dgc")

miseq

miseq_shallow_copy

miseq_deep_copy
```

You can see from the summaries that the dgc_data was added to both miseq and miseq_shallow_copy because they actually reference the same object, but miseq_deep_copy was not modified.
