---
title: "Functions for package developers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Functions for package developers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(rdataset)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

*rdataset* includes several functions designed for package developers. These functions provide additional access to the back end data and allow developers to merge, remove and set data directly. The function names begin with *'xdev_'* or *'xint_'* to clearly indicate that they are designed for development or for internal use, and not for the general user. 

## Setting functions

Over the course of an analysis your package may want to change the abundances of sequences or bins, modify sequence nucleotide strings due to alignment, screening or filtering, and change the dataset name. Let's take a close look at functions you will need to do so using the `miseq_sop_example()' dataset. 

```{r}
miseq <- miseq_sop_example()
miseq
```

### Setting the dataset name

To change the dataset name you can use the `xdev_set_dataset_name()` function.

```{r}
names(data = miseq, type = "dataset")
xdev_set_dataset_name(data = miseq, dataset_name = "modified_miseq")
names(data = miseq, type = "dataset")
```

### Setting nucleotide sequence strings

The miseq example dataset is already aligned, filtered and screened, but for the sake of example let's set the nucleotide strings of sample 'F3D0' to 'NNNN'. First we will get the names of the sequences in sample 'F3D0' and look at the original sequence strings.

```{r}
sample_f3d0_names <- xdev_get_by_sample(
  data = miseq,
  type = "sequence_names",
  samples = c("F3D0")
)
head(sample_f3d0_names[[1]], 5)
sample_f3d0_sequences <- xdev_get_by_sample(
  data = miseq,
  type = "sequences",
  samples = c("F3D0")
)
head(sample_f3d0_sequences[[1]], 5)
```

Now we will assign nucleotide strings of the sequences in sample 'F3D0' to 'NNNN':

```{r}
nnnn_neucleotides <- rep("NNNN", length(sample_f3d0_names[[1]]))
comments <- rep("example_set_sequences", length(sample_f3d0_names[[1]]))

xdev_set_sequences(
  data = miseq, sequence_names = sample_f3d0_names[[1]],
  sequences = nnnn_neucleotides, comments = comments
)

xdev_get_by_sample(data = miseq,
                   type = "sequences", samples = c("F3D0"))[[1]] |> head(5)
```


### Setting sequence abundances

Now that we know how to set sequence strings, let's learn how to set sequence abundances. In most cases you can use the `assign()` function to set sequence abundances, but there may be cases where you want more flexibility. The assign function requires abundances to be provided for each sequence. If you have parallelized code using the dataset, you may want to allow each thread to update a set of sequences. There are two functions to that allow you to do that, `xdev_set_abundances()` and `xdev_set_abundance()`. *xdev_set_abundances* is used with datasets that include samples, *xdev_set_abundance* is used for datasets without samples. Let's set the abundances of all sequences in sample F3D0 to 0:

```{r}
num_samples <- count(data = miseq, type = "samples")
new_abunds <- rep(list(rep(0, num_samples)), length(sample_f3d0_names[[1]]))
xdev_set_abundances(
  data = miseq,
  sequence_names = sample_f3d0_names[[1]],
  abundances = new_abunds
)
```


### Setting bin abundances
