---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The *rdataset* package stores the data associated with your microbial DNA analysis. This includes nucleotide sequences, abundance, sample and treatment assignments, taxonomic classifications, asv, otu and phylotype clusters, metadata, trees and various reports. It is designed to facilitate data analysis across multiple R packages with utility functions to read from [mothur](https://mothur.org), [qiime2](https://qiime2.org), [dada2](https://benjjneb.github.io/dada2/) and [phyloseq](https://www.bioconductor.org/packages/release/bioc/html/phyloseq.html).

## Installing rdataset

You can install the CRAN version with:

```{r, eval = FALSE}
install.packages("rdataset")
```

You can install the development version of *rdataset* from [GitHub](https://github.com/SchlossLab/rdataset) with:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("SchlossLab/rdataset")
```

```{r setup}
library(rdataset)
```

## Importing Data

* [Importing data from mothur](vignettes/Importing_from_mothur.html)
* [Importing data from qiime2](vignettes/Importing_from_qiime2.html)

