---
title: "Importing from phyloseq"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Importing from phyloseq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(strollur)
library(phyloseq)
```

`strollur` includes functionality for the reading and writing of phyloseq objects. To convert a phyloseq object to a strollur object, you need to run the `read_phyloseq()` function. 

```{r}
# Using the phyloseq example data
phylo_object <- readRDS(strollur_example("GlobalPatterns.RDS"))
rdata_object <- read_phyloseq(phylo_object)
rdata_object
```

Now that are phyloseq object is converted into a strollur object, we can utilize functions like `count()`, `abundance()`, and `names()` to inspect the data. 
```{r rdataset_functions}
count(rdata_object, "samples")
head(names(rdata_object, "sequences"))
head(abundance(rdata_object, "sequences"))
```

Furthermore, we can output strollur objects as phyloseq objects using the `write_phyloseq()` function.

```{r write_phyloseq} 
phyloseq_object <- write_phyloseq(rdata_object)
phyloseq_object


# With the miseq example data
miseq <- miseq_sop_example()
miseq_phyloseq <- write_phyloseq(miseq)
miseq_phyloseq
```

