---
title: "Importing from qiime2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Importing from qiime2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(rdataset)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Importing Data From Qiime2

*rdataset* includes the function `read_qiime2()` as well as several functions to read [qiime2](https://qiime2.org) output files individually. To create a dataset from the outputs of the [qiime2 moving-pictures example](https://amplicon-docs.qiime2.org/en/latest/tutorials/moving-pictures.html), run the following:

```{r}
qza_files <- c(
  rdataset_example("rep_seqs.qza"),
  rdataset_example("table.qza"),
  rdataset_example("taxonomy.qza"),
  rdataset_example("rooted-tree.qza")
)

data <- read_qiime2(
  qza = qza_files,
  metadata = rdataset_example("sample_metadata.tsv"),
  dataset_name = "qiime_moving_pictures"
)
```

To view a summary of data:

```{r}
data
```

To create a dataset and read the individual files, you can use the functions below. First let's create a dataset named my_data.

```{r}
my_data <- new_dataset(dataset_name = "my_data")
```

To decompress the individual qza files, you can use the `unpack_qiime2_artifact()`. unpack_qiime2_artifact will extract a folder containing the data, versioning and provenance information. Let's take a look at the rep_seqs.qza file first.

```{r}
fasta_artifact <- unpack_qiime2_artifact(qza = rdataset_example("rep_seqs.qza"))
```

fasta_artifact is list. To add the [FASTA](https://www.ncbi.nlm.nih.gov/genbank/fastaformat/) data to your dataset you can use the `read_fasta()` function:

```{r}
fasta_file <- file.path(
  getwd(), "rep_seqs", fasta_artifact$uuid, "data",
  "dna-sequences.fasta"
)
fasta_data <- read_fasta(fasta = fasta_file)
```

fasta_data is a data.frame containing sequence names, sequence nucleotide strings, and comments if provided. You can add the FASTA sequences to your dataset using the `add()` function:

```{r}
add(data = my_data, table = fasta_data, type = "sequences")
my_data
```

You can remove the fasta_artifact using the unlink function:

```{r}
unlink(file.path(getwd(), "rep_seqs"), recursive = TRUE)
```

Now that we have added the FASTA sequences, let's unpack the abundance data found in the table.qza file.

```{r}
abundance_artifact <- unpack_qiime2_artifact(
  qza =
    rdataset_example("table.qza")
)
```

To read the abundance data you can use the `rbiom::`read_biom()` function:

```{r}
biom_file <- file.path(
  getwd(), "table", abundance_artifact$uuid, "data",
  "feature-table.biom"
)
biom_table <- rbiom::read_biom(biom_file)
abundance_data <- data.frame(
  sequence_names = biom_table$otus[biom_table$counts$i],
  abundances = biom_table$counts$v,
  samples = biom_table$samples[biom_table$counts$j]
)
```

abundance_data is a data.frame containing sequence_names, samples, and abundances. You can add the sequence abundance data to your dataset as follows:

```{r}
assign(data = my_data, table = abundance_data, type = "sequence_abundance")
my_data
```

You can remove the abundance_artifact using the following:

```{r}
unlink(file.path(getwd(), "table"), recursive = TRUE)
```

Since we know that the sequences are in fact features, we can assign the sequences to bins to create asv clusters. 

```{r}
assign(
  data = my_data, table = fasta_data, type = "bins", bin_type = "asv",
  table_names = list(
    sequence_name = "sequence_names",
    bin_name = "sequence_names"
  )
)
```

Now that we have added the abundance data and assigned the sequences to bins, let's take a look at the taxonomic classifications provided in the taxonomy.qza file.

```{r}
taxonomy_artifact <- unpack_qiime2_artifact(qza = rdataset_example(
  "taxonomy.qza"
))
```

To read the sequence classification data you can use the `read_table()` function:

```{r}
taxonomy_file <- file.path(
  getwd(), "taxonomy", taxonomy_artifact$uuid, "data",
  "taxonomy.tsv"
)
taxonomy_table <- read.table(taxonomy_file,
  sep = "\t", header = TRUE,
  quote = ""
)
names(taxonomy_table) <- c("sequence_names", "taxonomies", "confidences")
```

To add the classification data to your dataset, you can use the `assign()` function.

```{r}
assign(data = my_data, table = taxonomy_table, type = "sequence_taxonomy")
```

You can remove the taxonomy_artifact using the following:

```{r}
unlink(file.path(getwd(), "taxonomy"), recursive = TRUE)
```

Now, lets add a tree that shows the relationships between the sequences (features).

```{r}
tree_artifact <- unpack_qiime2_artifact(
  qza =
    rdataset_example("rooted-tree.qza")
)
```

We can read the tree using the `ape::read.tree()` function.

```{r}
tree_file <- file.path(
  getwd(), "rooted-tree", tree_artifact$uuid, "data",
  "tree.nwk"
)
sequence_tree <- ape::read.tree(tree_file)
```

You can add the tree to your dataset as follows:

```{r}
my_data$add_sequence_tree(tree = sequence_tree)
```

You can remove the tree_artifact using the following:

```{r}
unlink(file.path(getwd(), "rooted-tree"), recursive = TRUE)
```

Lastly, let's read the metadata provided by qiime2 using the `read_qiime2_metadata()` function. 

```{r}
metadata <- read_qiime2_metadata(
  metadata =
    rdataset_example("sample_metadata.tsv")
)
```

To add the metadata to the my_data dataset, run the following:

```{r}
add(data = my_data, table = metadata, type = "metadata")
```

To view a summary of your imported dataset, 'my_data', run the following:

```{r}
my_data
```
