---
title: "Using_Dataset"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using_Dataset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The *rdataset* package stores the data associated with your microbial DNA analysis. This tutorial will familiarize you with some of the functions available in the *rdataset* package. If you haven't reviewed the "Getting Started" tuturial, we recommend you start there. First let's load the rdataset package.

```{r setup}
library(rdataset)
```

We can use the **miseq_sop_example()** function to create a dataset object from the [Miseq SOP Example](https://mothur.org/wiki/miseq_sop/). 

```{r}
miseq <- miseq_sop_example()

# To print a summary of the miseq dataset
miseq
```

## Accessing Data
The *rdataset* package includes several functions to access your data. 

**names()** - The names function is an extension of base R's names function. It allows you to get the name of your dataset or the names of the sequences, bins, samples, treatments and reports in your dataset.

```{r}
# To get the name of the dataset
names(data = miseq, type = "dataset")

# To get the names of the sequences
names(data = miseq, type = "sequences") |> head(n = 5)

# To get the names of the sequences present in sample 'F3D0'
names(data = miseq, type = "sequences", samples = c("F3D0")) |> head(n = 5)

# To get the names of the sequences unique to sample 'F3D0'
names(data = miseq, type = "sequences", samples = c("F3D0"), distinct = TRUE) |>
  head(n = 5)

# To get the names of the "otu" bins
names(data = miseq, type = "bins", bin_type = "otu") |> head(n = 5)

# To get the names of the "asv" bins
names(data = miseq, type = "bins", bin_type = "asv") |> head(n = 5)

# To get the names of the "phylotype" bins
names(data = miseq, type = "bins", bin_type = "phylotype") |> head(n = 5)

# To get the names of the "otu" bins that are unique to 'F3D0'
names(
  data = miseq,
  type = "bins", samples = c("F3D0"), distinct = TRUE
) |> head(n = 5)

# To get the names of the "otu" bins that include sequences from 'F3D0'
names(
  data = miseq,
  type = "bins", samples = c("F3D0"), distinct = FALSE
) |> head(n = 5)

# To get the names of the samples
names(data = miseq, type = "samples")

# To get the names of the treatments
names(data = miseq, type = "treatments")

# To get the names of the reports
names(data = miseq, type = "reports")
```

**count()** - Similair to names() the count() function allows you to get the number of sequences, samples, treatments or bins in your dataset.

```{r}
# To get the total number of sequences
count(data = miseq, type = "sequences")

# To get number of unique sequences
count(data = miseq, type = "sequences", distinct = TRUE)

# To get number of unique sequences from samples 'F3D0' and 'F3D1'
# Note these sequences will be present in both samples but may be
# be present in other samples as well
count(data = miseq, type = "sequences", samples = c("F3D0", "F3D1"))

# To get number of unique sequences exclusive to samples 'F3D0' and 'F3D1'
# Note sequences are present in both samples and NOT present in any other
# samples.
count(
  data = miseq, type = "sequences", samples = c("F3D0", "F3D1"),
  distinct = TRUE
)

# To get the number of samples in the dataset
count(data = miseq, type = "samples")

# To get the number of treatments in the dataset
count(data = miseq, type = "treatments")

# To get the number of "otu" bins in the dataset
count(data = miseq, type = "bins", bin_type = "otu")

# To get the number of "asv" bins in the dataset
count(data = miseq, type = "bins", bin_type = "asv")

# To get the number of "phylotype" bins in the dataset
count(data = miseq, type = "bins", bin_type = "phylotype")

# To get number of "otu" bins from samples 'F3D0' and 'F3D1'
# Note these bins will have sequences from both samples but there may be
# other samples present as well
count(
  data = miseq,
  type = "bins", bin_type = "otu", samples = c("F3D0", "F3D1")
)

# To get number of "otu" bins unique to samples 'F3D0' and 'F3D1'
# Note these bins will have sequences from both samples and NO other samples
# will be present in the bins.
count(
  data = miseq, type = "bins", bin_type = "otu",
  samples = c("F3D0", "F3D1"), distinct = TRUE
)
```

* **abundance()** - You can use the abundance() function for access the abundance data for sequences, bins, samples, and treatments.

```{r}
# To the total abundance for each sequence
abundance(data = miseq, type = "sequences") |> head(n = 5)

# To the total abundance for each sequence parsed by sample
abundance(data = miseq, type = "sequences", by_sample = TRUE) |> head(n = 5)

# To the total abundance for each "otu" bin
abundance(data = miseq, type = "bins", bin_type = "otu") |> head(n = 5)

# To the total abundance for each "otu" bin parsed by sample
abundance(
  data = miseq,
  type = "bins", bin_type = "otu", by_sample = TRUE
) |> head(n = 5)

# To the total abundance for each "asv" bin
abundance(data = miseq, type = "bins", bin_type = "asv") |> head(n = 5)

# To the total abundance for each "asv" bin parsed by sample
abundance(
  data = miseq,
  type = "bins", bin_type = "asv", by_sample = TRUE
) |> head(n = 5)

# To the total abundance for each sample
abundance(data = miseq, type = "samples") |> head(n = 5)

# To the total abundance for each treatment
abundance(data = miseq, type = "treatments") |> head(n = 5)
```
    
* **report()** - The report function can be used to access the various data reports associated with your dataset. These reports include: 'sequence_data', 'sequence_taxonomy', 'bin_taxonomy', 'sequence_scrap', 'bin_scrap', 'metadata' and 'references'. If you have added custom reports for alignment, contigs_assembly or chimeras, you can get those as well using the type name you provided in add_reports(). 
 
```{r}
# To get a report about the FASTA sequence data
report(data = miseq, type = "sequence_data") |> head(n = 5)

# To get a classification report about your sequence data
report(data = miseq, type = "sequence_taxonomy") |> head(n = 7)

# To get the consensus taxonomies assigned to your 'otu' bins
report(data = miseq, type = "bin_taxonomy", bin_type = "otu") |> head(n = 7)

# To get the consensus taxonomies assigned to your 'asv' bins
report(data = miseq, type = "bin_taxonomy", bin_type = "asv") |> head(n = 7)

# To get the consensus taxonomies assigned to your 'phylotype' bins
report(
  data = miseq,
  type = "bin_taxonomy", bin_type = "phylotype"
) |> head(n = 7)

# To get the scrapped sequences report
report(data = miseq, type = "sequence_scrap")

# To get the "otu" scrapped bins report
report(data = miseq, type = "bin_scrap", bin_type = "otu")

# To get the metadata associated with your dataset
report(data = miseq, type = "metadata") |> head(n = 7)

# To get the resources associated with your dataset
report(data = miseq, type = "references")

# To access the custom reports, first let's find the names of the reports
names(data = miseq, type = "reports")

# To access our contigs assemble report
report(data = miseq, type = "contigs_report") |> head(n = 7)
```    


* **General**
    * **get_bin_types** - Returns the types of bins in your dataset
    * **is_aligned** - Returns TRUE if sequence data is aligned

```{r}
# To find the types of bins in your dataset
get_bin_types(data = miseq)

# To determine if the FASTA data is aligned
is_aligned(data = miseq)
```

## Sample Trees and Sequence Trees
#TODO


