---
title: "Using_Dataset"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using_Dataset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The *rdataset* package stores the data associated with your microbial DNA analysis. This tutorial will familiarize you with some of the functions available in the *rdataset* package. If you haven't reviewed the "Getting Started" tuturial, we recommend you start there. First let's load the rdataset package.

```{r setup}
library(rdataset)
```

We can use the **miseq_sop_example** function to create a dataset object from the [Miseq SOP Example](https://mothur.org/wiki/miseq_sop/). 

```{r}
miseq <- miseq_sop_example()
miseq
```

## Sequence Data
The *rdataset* package includes several functions to access your sequence data. 

* **Abundance**
    * **get_sequence_abundances** - Returns the total abundance for each sequence
    * **get_sequence_abundances_by_sample** - Returns the abundances of each sequence in an parsed by sample.
    * **get_sequence_abundance_table** - Returns a data.frame containing the sequence abundance data
    
```{r}
# To get the total abundance of each sequence:
sequence_abundances <- get_sequence_abundances(miseq)
head(sequence_abundances, n = 10)

# To get the total abundance of each sequence parsed by sample
sequence_abundances_by_sample <- get_sequence_abundances_by_sample(miseq)
head(sequence_abundances_by_sample[[2]], n = 10)

# To a data.frame containing the abundance of each sequence parsed by sample
sequence_abundance_table <- get_sequence_abundance_table(miseq)
head(sequence_abundance_table, n = 10)
```
    
* **Fasta, Classifications**
    * **get_sequence_names** - Returns the names of the sequences
    * **get_sequence_names_by_sample** - Returns the names of the sequences parsed by sample
    * **get_sequences** - Returns the nucleotide strings for each sequence
    * **get_sequences_by_sample** - Returns the nucleotide strings for each sequence parsed by sample
    * **is_aligned** - Returns TRUE if sequence data is aligned
    * **has_sequence_strings** - Returns TRUE if your dataset contains sequence nucleotide strings
    * **num** - Find the number of sequences, samples, treatments or bins of a given type
    
```{r}
# To get all the names of the sequences
seq_names <- get_sequence_names(miseq)
head(seq_names, n = 10)

# To get the names of the sequences in sample 'F3D0'
seq_names_f3d0 <- get_sequence_names(miseq, "F3D0")
head(seq_names_f3d0, n = 10)

# To get the names of the sequences parsed by sample
seq_names <- get_sequence_names_by_sample(miseq)
head(seq_names[[10]], n = 10)

# To determine if your dataset contains nucleotide strings
has_sequence_strings(miseq)

# To get all of the sequence nucleotide strings
seqs <- get_sequences(miseq)
head(seqs, n = 10)

# To get nucleotide strings for the sequences in sample 'F3D0'
seqs_f3d0 <- get_sequences(miseq, "F3D0")
head(seqs_f3d0, n = 10)

# To get nucleotide strings parsed by sample
seqs <- get_sequences_by_sample(miseq)
head(seqs[[10]], n = 10)

# To determine if the sequence data is aligned
is_aligned(miseq)

# To get the sequence classifications
seqs_taxonomy <- report(miseq, "sequence_taxonomy")
head(seqs_taxonomy, n = 10)

# To find the total number of sequences:
num(data = miseq, type = "sequences")

# To find the number of unique sequences:
num(data = miseq, type = "sequences", distinct = TRUE)

# To find total number of sequences from a given sample:
num(data = miseq, type = "sequences", sample = "F3D0")

# To find the number of unique sequence from a given sample:
num(data = miseq, type = "sequences", distinct = TRUE, sample = "F3D0")
```    

* **Sample and Treatment Functions**
    * **get_samples** - Returns the names of the samples
    * **get_treatments** - Returns the names of the treatments
    * **totals** - Returns the total number of sequences assigned to each sample or treatment
    * **get_sample_treatment_assignments** - Return a data.frame with sample treatment assignments
    * **has_sample** - Returns TRUE if a given sample is present
    
```{r}
# To find the number of samples
num(data = miseq, type = "samples")

# To get the names of the samples
get_samples(data = miseq)

# To get the number of sequences in each sample
totals(data = miseq, type = "samples")

# To find the number of treatments
num(data = miseq, type = "treatments")

# To get the names of the treatments
get_treatments(data = miseq)

# To get the number of sequences in each treatment
totals(data = miseq, type = "treatments")

# To get a table with sample treatment assignments
get_sample_treatment_assignments(data = miseq)

# To determine if sample 'F3D0' is present
has_sample(data = miseq, sample = "F3D0")

# To determine if sample 'sample_not_in_dataset' is present
has_sample(data = miseq, sample = "sample_not_in_dataset")
```
    
## Bin Data
The *rdataset* package includes several functions to access your bin data.

* **Bin Abundance**
    * **get_bin_abundance** - Returns the total abundance of a given bin
    * **get_bin_abundances** - Returns the abundances of a given bin parsed by sample.
    
```{r}
# To get the abundance of a given otu
get_bin_abundance(miseq, "Otu001", "otu")

# To get the abundance of a given phylotype bin
get_bin_abundance(miseq, "Phylo01", "phylotype")

# To get the abundance of a given otu parsed by sample
get_bin_abundances(miseq, "Otu001", "otu")

# To get the abundance of a given phylotype bin parsed by sample
get_bin_abundances(miseq, "Phylo01", "phylotype")
```

* **Bin Assignments**
    * **get_bin_types** - Returns the types of bins in your dataset
    * **get_bin_names** - Returns the names of the bins
    * **get_bin_assignments** - Returns a data.frame containing bin_names, bin abundances, samples (if assigned), and treatments (if assigned)
    * **get_bin** - Returns the names of sequences in a given bin
    * **get_list** - Returns a data.frame assigning sequences to bins
    * **get_list_vector** - Returns vector of strings containing the names of the sequences in each bin separated by commas
    * **get_rabund** - Returns a data.frame assigning abundances to bins
    * **get_rabund_vector** - Returns vector containing each bins abundance
    * **get_shared_vector** - Returns 2D vector containing the abundances of each bin parsed by sample.
    
```{r}
# To find the types of bins in your dataset
get_bin_types(miseq)

# To find the number of 'otu' bins
num(data = miseq, type = "bins", bin_type = "otu")

# To find the number of 'asv' bins
num(data = miseq, type = "bins", bin_type = "asv")

# To find the number of 'phylotype' bins
num(data = miseq, type = "bins", bin_type = "phylotype")

# To get the names of the 'otu' bins
otu_names <- get_bin_names(miseq, "otu")
head(otu_names, n = 10)

# To get the bin assignment of the 'otu' bins
bin_assignments <- get_bin_assignments(miseq, "otu")
head(bin_assignments, n = 10)

# To get the names of sequences in the bin named "Otu005"
get_bin(miseq, "Otu005", "otu")

# To get the sequence bin assignments for the 'otu' bins
otu_list <- get_list(miseq, "otu")
head(otu_list, n = 10)

# To get the sequence bin assignments for the 'phylotype' bins
phylo_list <- get_list(miseq, "phylotype")
head(phylo_list, n = 10)

# To get the sequence bin assignments for the 'otu' bins in vector format
otu_list <- get_list_vector(miseq, "otu")

# To get the bin abundances for the 'otu' bins
otu_abundance <- get_rabund(miseq, "otu")
head(otu_abundance, n = 10)

# To get the bin abundances for the 'phylotype' bins
phylo_abundance <- get_rabund(miseq, "phylotype")
head(phylo_abundance, n = 10)

# To get the bin abundances for the 'otu' bins in vector format
otu_abundance <- get_rabund_vector(miseq, "otu")
head(otu_abundance, n = 10)

# To get the bin abundances for the 'otu' bins parsed by sample
otu_shared_abundance <- get_shared_vector(miseq, "otu")
head(otu_shared_abundance, n = 10)
```

    
* **Bin Classifications and Representatives**
    * **report** - Returns the classification assignments for your bins
    * **get_bin_representative_sequences** - Returns a data.frame containing the representative_names and representative_sequences
    
```{r}
# To get the consensus taxonomies assigned to your 'otu' bins
bin_tax <- report(miseq, "bin_taxonomy", "otu")
head(bin_tax, n = 10)

# To get the consensus taxonomies assigned to your 'asv' bins
bin_tax <- report(miseq, "bin_taxonomy", "asv")
head(bin_tax, n = 10)

# To get the consensus taxonomies assigned to your 'phylotype' bins
bin_tax <- report(miseq, "bin_taxonomy", "phylotype")
head(bin_tax, n = 10)

otu_bin_reps <- get_bin_representative_sequences(miseq, "otu")
head(otu_bin_reps, n = 10)
```

## Reports, Metadata and References

* **Reports**
    * **report** - Returns a data.frame containing the report requested
    * **get_sequence_summary** - Returns list of data.frames containing the summaries of all reports, and scrapped data
    
```{r}
# To get the FASTA sequence report
reports <- report(miseq, "sequence_data")
head(reports, n = 10)

# To get the scrapped sequences report
report(miseq, "sequence_scrap")

# To get the "otu" scrapped bins report
report(miseq, "bin_scrap", "otu")

# To get the metadata associated with your dataset
report(miseq, "metadata")

# To get the resources associated with your dataset
report(miseq, "references")

# To get the summary reports
summary <- get_sequence_summary(miseq)
head(summary[[1]], n = 10)
```

## Sample Trees and Sequence Trees
#TODO

## General
* **Utility Functions**
    * **get_dataset_name** - Return the name of your dataset
    
```{r}
# To get the dataset name
get_dataset_name(miseq)
```
