---
title: "Using Dataset"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using Dataset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The *rdataset* package stores the data associated with your microbial DNA analysis. This tutorial will familiarize you with some of the functions available in the *rdataset* package. If you haven't reviewed the "Getting Started" tuturial, we recommend you start there. 

First let's load the rdataset package.
```{r setup}
library(rdataset)
```

## Loading the example dataset

We can use the **miseq_sop_example()** function to create a dataset object from the [Miseq SOP Example](https://mothur.org/wiki/miseq_sop/). 

```{r}
miseq <- miseq_sop_example()
```

To print a summary of the miseq dataset, run the following:

```{r}
miseq
```

## Accessing Data
The *rdataset* package includes several functions to access your data. 

**names()** - The names() function is an extension of base R's names function. It allows you to get the name of your dataset or the names of the sequences, bins, samples, treatments and reports in your dataset.

**count()** - The count() function allows you to get the number of sequences, samples, treatments or bins in your dataset.

**abundance()** - The abundance() function can be used to access the abundance data for sequences, bins, samples, and treatments.

**report()** - The report() function can be used to access the various data reports associated with your dataset. If you have added custom reports for alignment, contigs_assembly or chimeras, you can get those as well using the report_type name you provided in the add() function. 

**summary()** - The summary() function allows you to summarize sequences, your custom reports, and scrapped data. 


### names() and count() 
Let's take a closer look at the names() and count() functions. 

* **Parameters:** 
    + *data* - a dataset object
    + *type* - string containing the type of data you would like.
    + *bin_type* - string containing the bin type you would like data for.
    + *samples* - vector of strings containing sample names. samples is only used when 'type' = "sequences" or 'type' = "bins" .
    + *distinct* - boolean. 'distinct' can be used when 'type' is 'sequences' or 'type' is 'bins'. When 'type' = 'sequences' and distinct is TRUE the number of unique sequences is returned. When 'type' = 'sequences' and distinct is FALSE the total number of sequences is returned. When 'type' = "bins", you can set distinct = TRUE to return the number of bins that ONLY contain sequences from the given samples. When distinct is FALSE the count returned contains bins with sequences from a given samples, but those bins may also contain other samples.

To get the name of the dataset, set the type parameter to 'dataset':
```{r}
names(data = miseq, type = "dataset")
```

To get the names of the sequences in your dataset, set the type parameter to 'sequences':
```{r}
names(data = miseq, type = "sequences") |> head(n = 5)
```

To find the number of sequences in your dataset, you can use the count() function. Set the type to 'sequences'. Take special note of the 'distinct' parameter.  

To find the total number of sequences in the dataset, run the following:
```{r}
count(data = miseq, type = "sequences", distinct = FALSE)
```

To find the number of unique sequences in the dataset, set 'distinct' to TRUE:
```{r}
count(data = miseq, type = "sequences", distinct = TRUE)
```

To get the names of the unique sequences, run the following:
```{r}
names(data = miseq, type = "sequences") |> head(n = 5)
```

To get number of sequences from sample 'F3D0', you can set the samples parameter.
Note, these sequences will be present in the sample but may be be present in other samples as well.

```{r}
count(data = miseq, type = "sequences", samples = c("F3D0"))
```

To get the names of the sequences present in sample 'F3D0'
```{r}
names(data = miseq, type = "sequences", samples = c("F3D0")) |> head(n = 5)
```

To get number of unique sequences exclusive to sample 'F3D0'
*Note sequences are present in the sample and NOT present in any other samples.*
```{r}    
count(data = miseq, type = "sequences", samples = c("F3D0"), distinct = TRUE)
```

To get the names of the sequences unique to sample 'F3D0'
```{r}
names(data = miseq, type = "sequences", samples = c("F3D0"), distinct = TRUE) |>
  head(n = 5)
```

To get the number of "otu" bins in the dataset
```{r}
count(data = miseq, type = "bins", bin_type = "otu")
```

To get the names of the "otu" bins
```{r}
names(data = miseq, type = "bins", bin_type = "otu") |> head(n = 5)
```

To get the number of "asv" bins in the dataset
```{r}
count(data = miseq, type = "bins", bin_type = "asv")
```

To get the names of the "asv" bins
```{r}
names(data = miseq, type = "bins", bin_type = "asv") |> head(n = 5)
```

To get the number of "phylotype" bins in the dataset
```{r}
count(data = miseq, type = "bins", bin_type = "phylotype")
```

To get the names of the "phylotype" bins
```{r}
names(data = miseq, type = "bins", bin_type = "phylotype") |> head(n = 5)
```

To get number of "otu" bins from sample 'F3D0'
*Note these bins will have sequences from the sample but there may be other*
*samples present in the bins as well*
```{r}
count(data = miseq, type = "bins", bin_type = "otu", samples = c("F3D0"))
```

To get the names of the "otu" bins that include sequences from 'F3D0'
```{r}
names(
  data = miseq,
  type = "bins", samples = c("F3D0"), distinct = FALSE
) |> head(n = 5)
```

To get number of "otu" bins unique to samples 'F3D0'
*Note these bins will have sequences from the sample and NO other samples will* 
*be present in the bins.*
```{r}
count(
  data = miseq, type = "bins",
  bin_type = "otu", samples = c("F3D0"), distinct = TRUE
)
```

To get the names of the "otu" bins that are unique to 'F3D0'
```{r}
names(
  data = miseq,
  type = "bins", samples = c("F3D0"), distinct = TRUE
) |> head(n = 5)
```

To get the number of samples in the dataset
```{r}
count(data = miseq, type = "samples")
```

To get the names of the samples
```{r}
names(data = miseq, type = "samples")
```

To get the number of treatments in the dataset
```{r}
count(data = miseq, type = "treatments")
```

To get the names of the treatments
```{r}
names(data = miseq, type = "treatments")
```

To get the names of the reports
```{r}
names(data = miseq, type = "reports")
```

Now that we are familar with the names() and count() functions. Let's take a closer look at how we can use the **abundance()** function.

```{r}
# To the total abundance for each sequence
abundance(data = miseq, type = "sequences") |> head(n = 5)

# To the total abundance for each sequence parsed by sample
abundance(data = miseq, type = "sequences", by_sample = TRUE) |> head(n = 5)

# To the total abundance for each "otu" bin
abundance(data = miseq, type = "bins", bin_type = "otu") |> head(n = 5)

# To the total abundance for each "otu" bin parsed by sample
abundance(
  data = miseq,
  type = "bins", bin_type = "otu", by_sample = TRUE
) |> head(n = 5)

# To the total abundance for each "asv" bin
abundance(data = miseq, type = "bins", bin_type = "asv") |> head(n = 5)

# To the total abundance for each "asv" bin parsed by sample
abundance(
  data = miseq,
  type = "bins", bin_type = "asv", by_sample = TRUE
) |> head(n = 5)

# To the total abundance for each sample
abundance(data = miseq, type = "samples") |> head(n = 5)

# To the total abundance for each treatment
abundance(data = miseq, type = "treatments") |> head(n = 5)
```
    
The **report()** and **summary()** functions allow you to access sequence and  classification reports, metadata, resource references, scrapped data reports, sequence data summaries, custom report summaries and scrapped data summaries. Let's take a closer look using the miseq example.
 
```{r}
# To get a report about the FASTA sequence data
report(data = miseq, type = "sequence_data") |> head(n = 5)

# To summarize the FASTA sequence data
summary(data = miseq, type = "sequences", verbose = FALSE)

# To get a classification report about your sequence data
report(data = miseq, type = "sequence_taxonomy") |> head(n = 7)

# To get the consensus taxonomies assigned to your 'otu' bins
report(data = miseq, type = "bin_taxonomy", bin_type = "otu") |> head(n = 7)

# To get the consensus taxonomies assigned to your 'asv' bins
report(data = miseq, type = "bin_taxonomy", bin_type = "asv") |> head(n = 7)

# To get the consensus taxonomies assigned to your 'phylotype' bins
report(data = miseq, type = "bin_taxonomy", bin_type = "phylotype") |>
  head(n = 7)

# To get the scrapped sequences report
# Note, the miseq example does not include scrapped data
report(data = miseq, type = "sequence_scrap")

# To get the "otu" scrapped bins report
# Note, the miseq example does not include scrapped data
report(data = miseq, type = "bin_scrap", bin_type = "otu")

# To get a summary of the data scrapped during your analysis
# Note, the miseq example does not include scrapped data
summary(data = miseq, type = "scrap", verbose = FALSE)

# To get the metadata associated with your dataset
report(data = miseq, type = "metadata") |> head(n = 7)

# To get the resource references associated with your dataset
report(data = miseq, type = "references")

# To access the custom reports, first let's find the names of the reports
names(data = miseq, type = "reports")

# To access our contigs assembly report
report(data = miseq, type = "contigs_report") |> head(n = 7)

# To get a summary of your contigs assembly report
summary(
  data = miseq, type = "reports",
  report_type = "contigs_report", verbose = FALSE
)
```    

## General

**copy_dataset()** - Create a new dataset from an existing dataset.

**save_dataset()** - Save a dataset as a .rds file.

**load_dataset()** - Create a new dataset from a .rds file.

**export_dataset()** - Create a human readable list containing the dataset data.

**import_dataset()** - Create a new dataset from an exported dataset.

**get_bin_types()** - Returns the types of bins in your dataset

**is_aligned()** - Returns TRUE if sequence data is aligned

```{r}
# To create a new dataset that is a copy of miseq
copy_of_miseq <- copy_dataset(data = miseq)

# To save the dataset as a .rds file
save_dataset(data = miseq, file = "miseq.rds")

# To create a dataset from a .rds file
data <- load_dataset(file = "miseq.rds")

# To export the data into a human readable format
table <- export_dataset(data = miseq)

# For an overview of the exported data.frames
names(table)

# To create a new dataset from the exported table
data <- import_dataset(table = table)

# To find the types of bins in your dataset
get_bin_types(data = miseq)

# To determine if the FASTA data is aligned
is_aligned(data = miseq)
```

## Sample Trees and Sequence Trees
#TODO


