---
title: "Using_Dataset"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using_Dataset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The *rdataset* package stores the data associated with your microbial DNA analysis. This tutorial will familiarize you with some of the functions available in the *rdataset* package. If you haven't reviewed the "Getting Started" tuturial, we recommend you start there. First let's load the rdataset package.

```{r setup}
library(rdataset)
```

We can use the **miseq_sop_example()** function to create a dataset object from the [Miseq SOP Example](https://mothur.org/wiki/miseq_sop/). 

```{r}
miseq <- miseq_sop_example()
miseq
```

## Accessing Data
The *rdataset* package includes several functions to access your data. 

**names()** - The names function is an extension of base R's names function. It allows you to get the name of your dataset or the names of the sequences, bins, samples, treatments and reports in your dataset.

```{r}
 # To get the name of the dataset
 names(data = miseq, type = "dataset")

 # To get the names of the sequences
 sequence_names <- names(data = miseq, type = "sequences")
 head(sequence_names, n = 5)

 # To get the names of the sequences present in sample 'F3D0'
 sequence_names <- names(data = miseq, type = "sequences", samples = c("F3D0"))
 head(sequence_names, n = 5)

# To get the names of the sequences unique to sample 'F3D0'
sequence_names <- names(
  data = miseq,
  type = "sequences",
  samples = c("F3D0"), distinct = TRUE
)
head(sequence_names, n = 5)

# To get the names of the samples
names(data = miseq, type = "samples")

# To get the names of the treatments
names(data = miseq, type = "treatments")

# To get the names of the "otu" bins
bin_names <- names(data = miseq, type = "bins", bin_type = "otu")
head(bin_names, n = 5)

# To get the names of the "asv" bins
bin_names <- names(data = miseq, type = "bins", bin_type = "asv")
head(bin_names, n = 5)

# To get the names of the "phylotype" bins
bin_names <- names(data = miseq, type = "bins", bin_type = "phylotype")
head(bin_names, n = 5)

# To get the names of the "otu" bins that are unique to 'F3D0'
bin_names <- names(
  data = miseq,
  type = "bins", samples = c("F3D0"), distinct = TRUE
)
head(bin_names, n = 5)

# To get the names of the "otu" bins that include sequences from 'F3D0'
bin_names <- names(
  data = miseq,
  type = "bins", samples = c("F3D0"), distinct = FALSE
)
head(bin_names, n = 5)

# To get the names of the reports
names(data = miseq, type = "reports")
```

**count()** - Similair to names() the count() function allows you to get the number of sequences, samples, treatments or bins in your dataset.

```{r}
# To get the total number of sequences
count(data = miseq, type = "sequences")

# To get number of unique sequences
count(data = miseq, type = "sequences", distinct = TRUE)

# To get number of unique sequences from samples 'F3D0' and 'F3D1'
# Note these sequences will be present in both samples but may be
# be present in other samples as well
count(data = miseq, type = "sequences", samples = c("F3D0", "F3D1"))

# To get number of unique sequences exclusive to samples 'F3D0' and 'F3D1'
# Note sequences are present in both samples and NOT present in any other
# samples.
count(
  data = miseq, type = "sequences", samples = c("F3D0", "F3D1"),
  distinct = TRUE
)

# To get the number of samples in the dataset
count(data = miseq, type = "samples")

# To get the number of treatments in the dataset
count(data = miseq, type = "treatments")

# To get the number of "otu" bins in the dataset
count(data = miseq, type = "bins", bin_type = "otu")

# To get the number of "asv" bins in the dataset
count(data = miseq, type = "bins", bin_type = "asv")

# To get the number of "phylotype" bins in the dataset
count(data = miseq, type = "bins", bin_type = "phylotype")

# To get number of "otu" bins from samples 'F3D0' and 'F3D1'
# Note these bins will have sequences from both samples but there may be
# other samples present as well
count(
  data = miseq,
  type = "bins", bin_type = "otu", samples = c("F3D0", "F3D1")
)

# To get number of "otu" bins unique to samples 'F3D0' and 'F3D1'
# Note these bins will have sequences from both samples and NO other samples
# will be present in the bins.
count(
  data = miseq, type = "bins", bin_type = "otu",
  samples = c("F3D0", "F3D1"), distinct = TRUE
)
```


* **abundance**

* **Fasta, Classifications**
    * **get_sequences** - Returns the nucleotide strings for each sequence
    * **is_aligned** - Returns TRUE if sequence data is aligned
    * **has_sequence_strings** - Returns TRUE if your dataset contains sequence nucleotide strings
    * **num** - Find the number of sequences, samples, treatments or bins of a given type
    
```{r}
# To determine if your dataset contains nucleotide strings
has_sequence_strings(miseq)

# To get all of the sequence nucleotide strings
seqs <- get_sequences(miseq)
head(seqs, n = 10)

# To get nucleotide strings for the sequences in sample 'F3D0'
seqs_f3d0 <- get_sequences(miseq, "F3D0")
head(seqs_f3d0, n = 10)

# To determine if the sequence data is aligned
is_aligned(miseq)

# To get the sequence classifications
seqs_taxonomy <- report(miseq, "sequence_taxonomy")
head(seqs_taxonomy, n = 10)
```    

* **Sample and Treatment Functions**
    * **totals** - Returns the total number of sequences assigned to each sample or treatment
    * **get_sample_treatment_assignments** - Return a data.frame with sample treatment assignments
    * **has_sample** - Returns TRUE if a given sample is present
    
```{r}
# To get the number of sequences in each sample
count(data = miseq, type = "samples")

# To get the number of sequences in each treatment
count(data = miseq, type = "treatments")

# To the total abundance for each sample
abundance(data = miseq, type = "samples")

# To the total abundance for each treatment
abundance(data = miseq, type = "treatments")

# To get a table with sample treatment assignments
get_sample_treatment_assignments(data = miseq)

# To determine if sample 'F3D0' is present
has_sample(data = miseq, sample = "F3D0")

# To determine if sample 'sample_not_in_dataset' is present
has_sample(data = miseq, sample = "sample_not_in_dataset")
```
    
## Bin Data
The *rdataset* package includes several functions to access your bin data.
    
```{r}
 # To the total abundance for each "otu" bin
 abundance(data = miseq, type = "bins", bin_type = "otu")

 # To the total abundance for each "otu" bin parsed by sample
 abundance(data = miseq, type = "bins", bin_type = "otu", by_sample = TRUE)

 # To the total abundance for each "asv" bin
 abundance(data = miseq, type = "bins", bin_type = "asv")

 # To the total abundance for each "asv" bin parsed by sample
 abundance(data = miseq, type = "bins", bin_type = "asv", by_sample = TRUE)
```

* **Bin Assignments**
    * **get_bin_types** - Returns the types of bins in your dataset
    * **get_bin_assignments** - Returns a data.frame containing bin_names, bin abundances, samples (if assigned), and treatments (if assigned)
    * **get_bin** - Returns the names of sequences in a given bin
    * **get_list** - Returns a data.frame assigning sequences to bins
    * **get_list_vector** - Returns vector of strings containing the names of the sequences in each bin separated by commas
    * **get_rabund** - Returns a data.frame assigning abundances to bins
    * **get_rabund_vector** - Returns vector containing each bins abundance
    * **get_shared_vector** - Returns 2D vector containing the abundances of each bin parsed by sample.
    
```{r}
# To find the types of bins in your dataset
get_bin_types(miseq)

# To get the bin assignment of the 'otu' bins
#bin_assignments <- get_bin_assignments(miseq, "otu")
#head(bin_assignments, n = 10)

# To get the names of sequences in the bin named "Otu005"
#get_bin(miseq, "Otu005", "otu")

# To get the sequence bin assignments for the 'otu' bins
otu_list <- get_list(miseq, "otu")
head(otu_list, n = 10)

# To get the sequence bin assignments for the 'phylotype' bins
phylo_list <- get_list(miseq, "phylotype")
head(phylo_list, n = 10)

# To get the sequence bin assignments for the 'otu' bins in vector format
otu_list <- get_list_vector(miseq, "otu")

# To get the bin abundances for the 'otu' bins
otu_abundance <- abundance(data = miseq, type = "bins", bin_type = "otu")
head(otu_abundance, n = 10)

# To get the bin abundances for the 'phylotype' bins
phylo_abundance <- abundance(data = miseq,
                             type = "bins", bin_type = "phylotype")
head(phylo_abundance, n = 10)

# To get the bin abundances for the 'otu' bins parsed by sample
otu_shared_abundance <- abundance(data = miseq,
                                  type = "bins",
                                  bin_type = "otu",
                                  by_sample = TRUE)
head(otu_shared_abundance, n = 10)
```

    
* **Bin Classifications and Representatives**
    * **report** - Returns the classification assignments for your bins
    * **get_bin_representative_sequences** - Returns a data.frame containing the representative_names and representative_sequences
    
```{r}
# To get the consensus taxonomies assigned to your 'otu' bins
bin_tax <- report(miseq, "bin_taxonomy", "otu")
head(bin_tax, n = 10)

# To get the consensus taxonomies assigned to your 'asv' bins
bin_tax <- report(miseq, "bin_taxonomy", "asv")
head(bin_tax, n = 10)

# To get the consensus taxonomies assigned to your 'phylotype' bins
bin_tax <- report(miseq, "bin_taxonomy", "phylotype")
head(bin_tax, n = 10)

otu_bin_reps <- get_bin_representative_sequences(miseq, "otu")
head(otu_bin_reps, n = 10)
```

## Reports, Metadata and References

* **Reports**
    * **report** - Returns a data.frame containing the report requested
    * **get_sequence_summary** - Returns list of data.frames containing the summaries of all reports, and scrapped data
    
```{r}
# To get the FASTA sequence report
reports <- report(miseq, "sequence_data")
head(reports, n = 10)

# To get the scrapped sequences report
report(miseq, "sequence_scrap")

# To get the "otu" scrapped bins report
report(miseq, "bin_scrap", "otu")

# To get the metadata associated with your dataset
report(miseq, "metadata")

# To get the resources associated with your dataset
report(miseq, "references")

# To get the summary reports
summary <- get_sequence_summary(miseq)
head(summary[[1]], n = 10)
```

## Sample Trees and Sequence Trees
#TODO

